R for Clinical Study Reports and Submission

### 2021-09-19 tlf-disposition.Rmd
address conflicts

Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf), a summary needs to be provided that accounts for all participants who entered the study within Section 10.1, Disposition of participants.
This section summarizes the number of participants that were randomized, entered and completed each phase of the study. In addition, the reasons for all post-randomization discontinuations, grouped by treatment and by major reason of discontinuation are required (lost to follow-up, adverse event, ...).

@@

```{r}
adsl <- haven::read_sas("adam_data/adsl.sas7bdat")
```

Following variables are used in the preparation of the disposition report:

@@

```{r}
adsl %>%
  dplyr::select(USUBJID, TRT01P, TRT01PN, DISCONFL, DCREASCD) %>%
  utils::head(4)
```

In the code below, the number of participants randomized is calculated.

```{r}
n_rand <- adsl %>%
  dplyr::group_by(TRT01P) %>%
  dplyr::summarize(n = n()) %>%
  tidyr::pivot_wider(
    names_from = TRT01P,
    names_prefix = "",
    values_from = n
  ) %>%
  dplyr::mutate(row = "participants Randomized")

@@

In the code below, the number of participants that discontinued the study is calculated.

@@

```{r}
n_disc <- adsl %>%
  dplyr::group_by(TRT01P) %>%
  dplyr::summarize(
    n = base::sum(DISCONFL == "Y"),
    pct = base::formatC(n / n() * 100,
      digits = 1, format = "f", width = 5
    )
  ) %>%
  dplyr:: mutate(npct = base::paste0(n, " (", pct, "%)")) %>%
  dplyr::select(-c(n,pct)) %>%
  tidyr::pivot_wider(
    names_from = TRT01P,
    values_from = npct,
    values_fill = list(npct = "0")
  ) %>%
  dplyr::mutate(row = "Discontinued")

@@

To calculate the number and percentage of participants that completed/discontinued the study for each discontinuation reason, below code can be used.

```{r}
n_reason <- adsl %>%
  dplyr::group_by(TRT01P) %>%
  dplyr::mutate(n_total = dplyr::n()) %>%
  dplyr::group_by(TRT01P, DCREASCD) %>%
  dplyr::summarize(
    n = dplyr::n(),
    pct = base::formatC(n / unique(n_total) * 100,
      digits = 1, format = "f", width = 5
    )
  ) %>%
  dplyr:: mutate(npct = base::paste0(n, " (", pct, "%)")) %>%
  tidyr::pivot_wider(
    id_cols = DCREASCD,
    names_from = TRT01P,
    values_from = npct,
    values_fill = list(npct = "0")
  ) %>%
  dplyr::rename(row = DCREASCD)

@@

From the step above, the number and percentage of participants that completed the study can be isolated so change the
roworder in the final table.

```{r}
n_complete <- n_reason %>% 
  dplyr::filter(row == "Completed")

@@

For display purposes leading spaces are added to study discontinuation. This will mimick identation in the final report.

```{r}
n_reason <- n_reason %>%
  dplyr::filter(row != "Completed") %>%
  dplyr::mutate(row = paste0("    ", row))

@@

```{r}
tbl_disp <- base::rbind(n_rand, n_complete, n_disc, n_reason)

@@

  # Table title
  rtf_title("Disposition of participants") %>%

>>>>

Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf), 
we need to summarize the accounting of all participants who entered the study in Section 10.1, Disposition of Participants.

The disposition of participant reports the numbers of participants who were randomized, 
and who entered and completed each phase of the study. 
In addition, the reasons for all post-randomization discontinuations, 
grouped by treatment and by major reason (lost to follow-up, adverse event, poor compliance, etc.).

@@

```{r}
adsl <- read_sas("adam_data/adsl.sas7bdat")
```

Following variables are used in the preparation of a simplified disposition of participants table:

@@

```{r}
adsl %>%
  select(USUBJID, TRT01P, TRT01PN, DISCONFL, DCREASCD) %>%
  head(4)
```

In the code below, we calculated participants in the analysis population.

```{r}
n_rand <- adsl %>%
  group_by(TRT01P) %>%
  summarize(n = n()) %>%
  pivot_wider(
    names_from = TRT01P,
    names_prefix = "",
    values_from = n
  ) %>%
  mutate(row = "Participants Randomized")

@@

In the code below, we calculate the number and percentage of participants who discontinued the study.

@@

```{r}
n_disc <- adsl %>%
  group_by(TRT01P) %>%
  summarize(
    n = sum(DISCONFL == "Y"),
    pct = formatC(n / n() * 100,
      digits = 1, format = "f", width = 5
    )
  ) %>%
  mutate(npct = paste0(n, " (", pct, "%)")) %>%
  select(-c(n,pct)) %>%
  pivot_wider(
    names_from = TRT01P,
    values_from = npct,
    values_fill = list(npct = "0")
  ) %>%
  mutate(row = "Discontinued")

@@

In the code below, we calculate the number and percentage of participants who completed/discontinued the study for different reasons.

```{r}
n_reason <- adsl %>%
  group_by(TRT01P) %>%
  mutate(n_total = n()) %>%
  group_by(TRT01P, DCREASCD) %>%
  summarize(
    n = n(),
    pct = formatC(n / unique(n_total) * 100,
      digits = 1, format = "f", width = 5
    )
  ) %>%
  mutate(npct = paste0(n, " (", pct, "%)")) %>%
  pivot_wider(
    id_cols = DCREASCD,
    names_from = TRT01P,
    values_from = npct,
    values_fill = list(npct = "0")
  ) %>%
  rename(row = DCREASCD)

@@

In the code below, we calculate the number and percentage of participants who complete the study.
We split `n_reason` because we need to customize the row order of the table.

```{r}
n_complete <- n_reason %>% 
  filter(row == "Completed")

@@

In the code below, we calculate the number and percentage of participants who discontinued the study for different reasons.
For display purposes, we use `paste0("    ", row)` to add leading spaces to mimick identation in the final report.

```{r}
n_reason <- n_reason %>%
  filter(row != "Completed") %>%
  mutate(row = paste0("    ", row))

@@

```{r}
tbl_disp <- rbind(n_rand, n_complete, n_disc, n_reason)

@@

  # Table title
  rtf_title("Disposition of Participants") %>%
